---
title: "Intro to Tidyverse"
author: "Jeremy Fan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: default
  pdf_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "")
```


 Challenge 2 code here(x <- gapminder%>%filter(year>=1980)%>%arrange((year))
 
 
##echo: (TRUE; logical or numeric) Whether to display the source code in the output document.

##message: whether to preserve the message emitted by message() in the source code

##warning: to perserve warnings or not

##prefix to be added before each line of the text output.. only relevant if you have text output, default is set to ##, but  making it "" will make it nothing


***
> These materials have been taken from the Software Carpentry: R Novice Lesson.
You can find the original materials
[here](https://github.com/swcarpentry/r-novice-gapminder)

This lesson will cover some basic functions that can be used to manipulate data
in R. Again, we will be using the gapminder data set, which includes country
information on GDP, population, etc.

There are five main functions we'll be talking about today, each allowing us to
manipulate data frames. These five functions are:

* `select()`  --  Choose columns (variables or attributes) from our data frame
* `filter()`  --  Choose rows (samples or observations) from our data frame
* `mutate()`  --  Create new columns, based on existing ones
* `group_by()`  --  Group rows based on a particular column/value within that column 
* `summarize()`  --  Perform some function on the grouped data

If you haven't already, make sure you have `dplyr()` and `gapminder()` installed
and loaded with the following commands:

```{r}
# Download the packages


# install.packages(c("tidyverse", "gapminder"))
# Load the packages for use
library(tidyverse)
library(gapminder)
#gapminder <- read.csv("../data/gapminder_data.csv")

#Alternatively 

if (!require('tidyverse')) install.packages('tidyverse'); library('tidyverse')
if (!require('gapminder')) install.packages('gapminder'); library('gapminder')
```

Let's take a quick look at our data frame to remind ourselves of its structure.
We do this using the `head()` command, which will display the first 6 rows
of our data frame.
```{r}
head(gapminder)
```

## Quick aside on tibbles and data frames
In R, one of the main types of objects/variables we're going to be working with
is a **data frame**. This is much like a table you would view in Excel, where
column represent variables or measures and rows represent measurements, samples,
or observations. 

Another data set is that a **tibble** which has 2 main differences. Tibbles never changes the input type - i.e No more worry of characters being automatically turned into strings. A tibble can have columns that are lists. There are other slight differences, for instance, tibbles can have non-standard variable names.

What are we going to be working with? 

```{r}
class(gapminder)
#The dataset that we are working with was originally a data.frame that has been coerced into a tibble, which is why multiple class names are returned by the `class()` function.
```

Gapminder is a df that was coerced into a tibble. 


## Choose Columns: select

The first function we'll be using is `select()`. This function let's us pick
columns from our data frame, based on name (e.g. year) or by index (e.g. 3).


Let's try using `select()` to pick out a few columns: "country", "year",
"lifeExp", and "pop". We'll be assigning these columns to a new data frame,
`gapminder_select`. Then we'll use `head()` to see if it worked.


```{r}

# select() code here
## three ways to select
(gapminder_select <- select(gapminder, country, year, lifeExp, pop))
(gapminder_remove <- select(gapminder, -continent, -gdpPercap))
(gapminder_select_index <- select(gapminder, 1,3:5)) #note we start indexing at 1


(gapminder_reorder <- select(gapminder, pop, everything())) #basically reorder a dataframe

(gapminder_rename <- rename(gapminder, population = pop )) #note that the current column name is the second value


gapminder_letter <- select(gapminder, starts_with("p"))  #can also choose starting letter

head(gapminder_rename)
```

As you can see, our new data frame contains only a subset of the columns from
the original data frame, based on the names we provided in the `select()`
command.


Here we'll also introduce another great feature of `dplyr()`: the pipe (
**%>%** ). This symbol sends or pipes an object (e.g. a data frame like
gapminder) INTO a function (e.g. `select()`).

So, the above `select()` command can be rewritten as follows (NOTE: the "." is a
placeholder, which represents the object being piped). Again, we can check our
result using `head()`.

```{r}
# select() using pipe syntax
gapminder_pipe <- gapminder %>% 
  select(., country, year, lifeExp, pop)


head(gapminder_pipe)
head(gapminder_select)
```


We can actually simplify the above command further - dplyr's functions such as
`select()` are smart enough that you don't actually need to include the "."
placeholder, as shown below.


```{r}
# select() using pipe syntax w/out a placeholder
gapminder_pipe2 <- gapminder %>% select(country, year, lifeExp, pop)
head(gapminder_pipe2)
```


### Challenge 1
Using the `select()` command and pipe (` %>% `) notation, pick the following
columns from the `gapminder` data frame, assign them to a new variable (we'll
use **x**), and display the results using `head(x, n = 10)`. Columns to choose
are:

* continent
* GDP per capita
* life expectancy
* year

```{r, eval = FALSE}
# Answer here:
(x <- gapminder %>%
  select(country, year, gdpPercap, lifeExp))

head(x, n = 10L)
```


## Choose Rows: filter

So we've covered selecting columns, but what about rows? This is where
`filter()` comes in. This function allows us to choose rows from our data frame
using some logical criteria. An example is filtering for rows in which the
country is Canada. This can also be applied to numerical values, such as the
year being equal to 1967, or life expectancy greater than 30.

NOTE: In R, equality (e.g. country is Canada, year is 1967) is done using a
double equals sign (`==`).


Let's go through a couple examples. 

```{r}
# Filter rows where country is Canada
gapminder_canada <- gapminder %>% filter(country == "Canada")
head(gapminder_canada)

## what if we wanted Canada and Albania?
gapminder_canada_albania <- gapminder %>%
    filter(country  %in%  c("Canada", "Albania"))
```

Let's try another one, this time filtering on life expectancy above a certain
threshold:

```{r}
# Filter for rows where life expectancy is greater than 50
gapminder_LE <- gapminder %>% filter(lifeExp > 50)
head(gapminder_LE)
```


We can also filter with multiple arguments, each separated by a comma:

```{r}
# filter() for Canada and life expectancy greater than 80
gapminder_C_LE <- gapminder %>% filter(country == "Canada", lifeExp > 80)
head(gapminder_C_LE)
```


### Challenge 2
Use `filter()` to choose data for African countries, from the year 1980 and
onwards.

```{r, eval = FALSE}
# Challenge 2 code here

```{r}
(x <- gapminder %>%
  filter(continent == "Africa", year >= 1980) %>%
  arrange(desc(year))) #default is asccending, if you want descending you ahve to do arrange(desc(year))

```


## Create New Columns: mutate()

Let's say we now want to calculate the GDP in billions, which is done by
mutiplying the GDP per capita by the population, then dividing by 1 billion (1 *
10^9). `mutate()` will perform this calculation on each row in the data frame,
one row at a time (i.e. row-wise). The code below will calculate the GDP in
billions:

* `gdpPercap * pop / 10^9`

```{r}
# Use mutate() to calculate GDP in billions
gapminder_gdpBil <- gapminder %>% 
  mutate(gdp_billion = gdpPercap * pop / 10^9)

head(gapminder_gdpBil)
```

```{r}
# Answer here:
(gapminder %>%
 select(country, continent, year) %>%
 mutate(across(everything(),as.factor)))
```
`
We can also model "growth" from the original 
```{r}
answer1.1 <-
    gapminder %>% 
    group_by(country) %>% 
    arrange(year) %>%
    mutate(rel_growth = pop - dplyr::first(pop))
```

## Combine Functions with Pipes
We've seen that pipes ( **%>%** ) can be used to send an object such as a data
frame into a function, such as `select()`, or `filter()`. But they can also be
used to send the output of one function into another function. This allows us to
chain together multiple commmands, without the need for intermediate variables.

Let's take a look at this in an example. 

```{r}
# select() the five columns, and filter() for Canada
gapminder_multi <- gapminder %>% 
    select(country, year, lifeExp, pop, gdpPercap) %>% 
    filter(country == "Canada")
head(gapminder_multi)
```


We can further expand on this by incorporating our `mutate()` command from
earlier, linking multiple functions into a single command. Be sure to indent
(`TAB` key) when moving to a new line after a pipe.

```{r}
# select() the four columns, filter() for Canada, and calculate GDP in billions
gapminder_multi_2 <- gapminder %>% 
    select(country, year, lifeExp, pop, gdpPercap) %>% 
    filter(country == "Canada") %>% 
    mutate(gdp_billion = gdpPercap * pop / 10^9)
head(gapminder_multi_2)
```


Lets say now we wanteed to actually create a categorical variable based on life expectancy in 1997.

```{r}
(gapminder_1997 <- gapminder %>% 
   filter(year == 1997) %>% 
   mutate(life_level = case_when(lifeExp < 23 ~ "very low",
                                 lifeExp < 48 ~ "low",
                                 lifeExp < 59 ~ "moderate",
                                 lifeExp < 70 ~ "high",
                                 TRUE ~ "very high")))
```

Are there any problems? COPY AND PASTE THIS CODE. Don't have to follow this one.. Just for demonstration:

```{r}
ggplot(gapminder_1997) + geom_boxplot(aes(x = life_level, y = gdpPercap)) +
  labs(y = "GDP per capita ($)", x= "Life expectancy level (years)") +
  ggtitle("GDP per capita per Level of Life Expectancy") +
  theme_bw() 
```

Two problems: very low had 0 samples, but shouldn't it be displayed on the graph. But more importantly, what's up with the ordering? high -> low -> moderate? Its alphabetically ordering it. 

To fix it, we need to establish levels: 

```{r}
(gapminder_1997_fct <- gapminder %>% 
   filter(year == 1997) %>% 
   mutate(life_level = factor(case_when(lifeExp < 23 ~ "very low", #note we added a factor
                                 lifeExp < 48 ~ "low",
                                 lifeExp < 59 ~ "moderate",
                                 lifeExp < 70 ~ "high",
                                 TRUE ~ "very high"),
                              levels = c("very low", "low", "moderate", "high", "very high")))) #note we add levels

```

```{r}
ggplot(gapminder_1997_fct) + geom_boxplot(aes(x = life_level, y = gdpPercap)) +
  labs(y = "GDP per capita ($)", x= "Life expectancy level (years)") +
  ggtitle("GDP per capita per Level of Life Expectancy") +
  theme_bw() 
```

## Calculations with group_by and summarise

These functions allow us to work on our data in specific groups. For example, we
can use `group_by()` to group observations by country, then calculate the
average life expectancy for each country. Summarise creates a new dataframe that will have on or more rows for each combination of grouping variables.

```{r}
# group_by() country, calculate average life expectancy
gapminder_grp <- gapminder %>% 
  group_by(country) %>% 
  summarise(mean(lifeExp))

head(gapminder_grp)
```


Let's do another example, again grouping by country. This time, we'll calculate
the mean and standard deviation of the GDP per capita. We'll also specify the
column names inside of the `summarise()` command.

```{r}
gapminder_mean_sd <- gapminder %>% 
    group_by(country) %>% 
    summarise(mean_gdp = mean(gdpPercap), sd_gdp = sd(gdpPercap))
head(gapminder_mean_sd)
```


## Tying it all together

Now let's use all the commands we've covered and combine them with pipes into a
single statement.

Let's say we want calculate the mean and SD of the GDP (in billions) for each
country, but only considering data from 1980 and onwards. We can accomplish this
all in one step as follows.

```{r}
# select() columns, filter() by year, calculate GDP in billions, mean() and sd()
# of GDP in billions
gapminder_final <- gapminder %>% 
    select(country, year, pop, gdpPercap) %>% 
    filter(year >= 1980) %>% 
    mutate(gdp_billion = gdpPercap * pop / 10^9) %>% 
    group_by(country) %>% 
    summarise(mean_gdpBillion = mean(gdp_billion), sd_gdpBillion = sd(gdp_billion))
head(gapminder_final)
```

Second final challenge
+ Add a column called `min_life` of the *minimum* life expectancy for each continent and each year of the `gapminder` dataset -- but keep the same amount of rows as in the original gapminder data set.
+ Add a column called `min_country` containing the country name having the lowest life expectancy for that year and that continent -- again, keeping the same amount of rows as the original.
+ `arrange` by min life expectancy

```{r}
(gapminder_final_2 <- gapminder %>% 
  group_by(continent) %>% 
  mutate(min_life = min(lifeExp),
               min_country = country[lifeExp == min_life]) %>%
  arrange(min_life))
```
