---
title: "ggplot Tips and Alternate Plotting Libraries"
author: "Travis Blimkie"
date: "October 2nd, 2019"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

***

Today we're going to look at a few ways of imporving on default ggplot2
graphics, as well as explore some other packages for plotting data.
Specifically, we're going to look at the `cowplot` package for tweaking our
ggplot2 figures, then produce interactive graphics via the `plotly` package.
Finally we'll use the built-in `mtcars` data set to test out these different
plotting methods.

Let's start by loading the libraries we'll need.  

```{r libraries}
# install.packages(c("cowplot", "plotly"))
library(cowplot)
library(plotly)
library(tidyverse)

data("mtcars")
mtcars <- mtcars %>% rownames_to_column("car")
```

<br>

Let's start by making a standard plot using `ggplot2` to get our baseline. 

```{r basic_ggplot}
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point()
```

<br>

Let's see how we can tweak this plot using the available methods. We'll be
making use of two functions here: `labs()` to create labels, and `theme()` to
customize the appearance of our plot. Let's change the axis labels, add a title,
and change the overall theme with `theme_bw()`.
```{r tweak_ggplot_1}
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point() +
  labs(x = "Number of Cylinders", 
       y = "Miles per Gallon", 
       title = "Comparing Engine Size and Fuel Efficiency") +
  theme_bw() +
  theme(axis.ticks = element_blank())
```

<br>

The `theme_bw()` function is acting as a "preset", changing multiple arguments
of the regular `theme()` function at once. There are some additional versions of
`theme_bw()`, such as `theme_classic()`, and `theme_grey()`. For `theme()`
itself, there are a whole range of arguments that can be used to alter the
appearance of the plot. Any argument you want to specify must be in the format
of `element_x()` (like the `element_blank()` we used earlier). Let's build on
this example by increasing the font size, as the default is often too small.
```{r tweak_ggplot_2}
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point(size = 3, pch = 21, fill = "dodgerblue4", colour = "black") +
  labs(x = "Number of Cylinders", 
       y = "Miles per Gallon", 
       title = "Comparing Engine Size and Fuel Efficiency") +
  theme_bw() +
  theme(axis.ticks = element_blank(),
        axis.text = element_text(size = 12, colour = "black"),
        title = element_text(size = 14, colour = "dodgerblue4"),
        panel.grid.minor = element_blank())
```

<br>

Now our plot looks a lot nicer!  

<br>

Another alternative to making a bunch of little changes ourselves is to use
another package, `cowplot`. This package comes with suite of functions, but
let's start with `theme_cowplot()`, which makes a variety of changes to the
default look of ggplot2 graphics.
```{r cowplot_1}
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point(size = 3, pch = 21, fill = "dodgerblue4", colour = "black") +
  labs(x = "Number of Cylinders", 
       y = "Miles per Gallon", 
       title = "Comparing Engine Size and Fuel Efficiency") +
  theme_cowplot()
```

<br>

Another really useful feature of the `cowplot` package is the ability to combine
multiple plots into a multi-panel style figure using the `plot_grid()` function.
Let's test it out with an example.
```{r plot_grid, fig.height = 8}
mpgCyl_plot <- ggplot(mtcars, aes(cyl, mpg)) +
  geom_point(size = 3, pch = 21, fill = "dodgerblue4", colour = "black") +
  labs(x = "Number of Cylinders", 
       y = "Miles per Gallon", 
       title = "Comparing Engine Size and Fuel Efficiency") +
  theme_cowplot()

dispHP_plot <- ggplot(mtcars, aes(disp, hp)) +
  geom_smooth(method = "lm", colour = "grey") +
  geom_point(size = 3, pch = 21, fill = "tomato", colour = "black") +
  labs(x = "Displacement", 
       y = "Horse Power", 
       title = "Comparing Engine Size and Displacement") +
  theme_cowplot()

plot_grid(mpgCyl_plot, dispHP_plot, nrow = 2, ncol = 1, align = "h", labels = "AUTO")
```

<br>

Now let's move on to another plotting library, `plotly`. Unlike ggplot2, this
package allows us to make interactive HTML plots. Let's see what this looks like
with an example. Note the syntax is a bit different then ggplot2, so pay close
attention to how we set up our plot.
```{r plotly_1}
plot_ly(mtcars, x = ~cyl, y = ~mpg, type = "scatter", text = ~car) %>% 
  plotly::layout(
        title = "Comparing Engine Size and Fuel Efficiency",
        xaxis = list(title = "Number of Cylinders"),
        yaxis = list(title = "Miles per Gallons")
  	)
```

<br>

As you can see, hovering our cursor over the points displays the coordinates as well as the name of the vehicle.  
Let's see how we can tweak this some more. 
```{r plotly_2}
fit <- lm(mpg ~ cyl, data = mtcars)

plot_ly(
  mtcars, x = ~cyl, y = ~mpg, type = "scatter", text = ~car
) %>% 
  add_lines(x = ~cyl, y = fitted(fit), line = list(color = "black")) %>% 
  plotly::layout(
    title = "Comparing Engine Size and Fuel Efficiency",
    xaxis = list(title = "Number of Cylinders"),
    yaxis = list(title = "Miles per Gallons")
  )
```

<br>

We can also extensively format the hover text displayed on the plot, using a bit
of HTML language.
```{r plotly_3}
fit <- lm(mpg ~ cyl, data = mtcars)

plot_ly(
  mtcars, 
  x = ~cyl
) %>%
  add_trace(y = ~mpg, 
            mode = "markers", 
            marker = list(size = 10),
            hoverinfo = "text",
            text = ~paste0("<b>Car: </b>", car)) %>% 
  add_trace(y = fitted(fit), mode = "lines") %>% 
  plotly::layout(
    title = "Comparing Engine Size and Fuel Efficiency",
    xaxis = list(title = "Number of Cylinders"),
    yaxis = list(title = "Miles per Gallons")
  )
```

<br>

Making a boxplot using plotly, specifying custom colours, and adding lines along
the x and y axis.
```{r plotly_4}
plot_ly(
  mtcars, 
  x = ~cyl
) %>%
  add_boxplot(
    y = ~mpg, 
    marker = list(color = "black"),
    line = list(color = "black"), 
    fillcolor = "white"
  ) %>% 
  plotly::layout(
    title = "Comparing Engine Size and Fuel Efficiency",
    xaxis = list(showline = TRUE, 
                 title = "Number of Cylinders", 
                 titlefont = list(size = 14), 
                 tickfont = list(size = 14)),
    yaxis = list(showline = TRUE, 
                 title = "Miles per Gallons",
                 titlefont = list(size = 14), 
                 tickfont = list(size = 14))
  )
```

<br>
