---
title: "Introduction to Leaflet"
author: "Jasmine"
date: "24/02/2020"
output: html_document
---

## Install `leaflet` and and other packages
```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("leaflet", "tidyverse", "readxl")
```

## Load libraries
```{r libraries}
library(tidyverse)
library(readxl)
library(leaflet)
```

## Read in the data
```{r load data, warning=FALSE}
ak <- read_xlsx("data/2019Sep26_algae_alaska.xlsx")
```
## Take a look at the data
- notice that there is the convenient `Geo_LatDecimal` and `Geo_LongDecimal`

### Important GIS terms
####`Datum` 
- *a set of reference points, used for locating places on the Earth*
- why so many? - the earth is not a perfect sphere -> different datums are better for different areas in the world (ie NAD84 for North America)
- usually it is in `WGS84` (ie google maps)
- important in GIS data because all coordinates must have the same Datum (but often not reported)
- mismatching your base map and your coordinate datum = weird results

#### `UTM` 
- Universal Transverse Mercator
- *a plane coordinate grid system named for the map projection*
- can also be converted to coordinate data
```{r}
glimpse(ak)
```

## Goal: Map one of the species

### 1. Filter the data frame
```{r}
df_ss <- ak %>% 
  filter(Genus == "Saccharina", Species ==  "subsimplex")
```

### 2. Map on to leaflet
#### Leaflet Basics
- similar to `ggplot` philosophy -> layering elements
- add commands start with `add____()`
- chain together layers by using `%>%`

#### start with basic leaflet call
```{r}
leaflet()
```

#### Basemap
- `?addProviderTiles` to see full list of providers
- Preview basemap tiles [here](https://leaflet-extras.github.io/leaflet-providers/preview/index.html)

```{r}
leaflet() %>% 
  addProviderTiles("OpenStreetMap.Mapnik") 
```

### Add some points
- this is all you need to make a basic map
- you don't need to find a base map or decide on an extent before hand
- good for large data sets
```{r}
leaflet() %>% 
  addProviderTiles("OpenStreetMap.Mapnik") %>% 
  addCircleMarkers(lat = df_ss$Geo_LatDecimal, lng = df_ss$Geo_LongDecimal)
```

## Now for some extras

### Add some different points
- `addAwesomeMarkers` gives you map pin markers
```{r}
leaflet() %>% 
  addProviderTiles("OpenStreetMap.Mapnik") %>% 
  addAwesomeMarkers(lat = df_ss$Geo_LatDecimal, lng = df_ss$Geo_LongDecimal)
```

### Add some information with the points
- now you can click on the points for a pop up bubble
```{r}
leaflet() %>% 
  addProviderTiles("OpenStreetMap.Mapnik") %>% 
  addCircleMarkers(lat = df_ss$Geo_LatDecimal, lng = df_ss$Geo_LongDecimal,
                   popup = paste("Collector:", df_ss$`Primary Collector`,
                                 "Date collected:", df_ss$`Date Collected`))
```

### Clustering
- using `markerClusterOptions()`
- useful for large datasets
```{r markerClusterOptions}
leaflet() %>% 
  addProviderTiles("OpenStreetMap.Mapnik") %>% 
  addCircleMarkers(lat = df_ss$Geo_LatDecimal, lng = df_ss$Geo_LongDecimal,
                   popup = paste("Collector:", df_ss$`Primary Collector`,
                                 "Date collected:", df_ss$`Date Collected`,
                                 ", Location:", df_ss$Location, 
                                 ", Collect no.:",df_ss$`Collector Number`),
                   clusterOptions = markerClusterOptions())
```

### Add buttons
- add zoom button using javascript (or use google)
```{r}
leaflet() %>% 
  addProviderTiles("OpenStreetMap.Mapnik") %>% 
  addCircleMarkers(lat = df_ss$Geo_LatDecimal, lng = df_ss$Geo_LongDecimal) %>% 
  addEasyButton(easyButton(icon = "fa-crosshairs", title = "Locate Me",
                           onClick = JS("function(btn, map){ map.setZoom(2);}")))
```


## Now try this with the BC data
```{r bc solutions, warning=FALSE}
bc <- read_xlsx("data/2019Sep26_algae_bc.xlsx")

top_sp_bc <- bc %>% 
  mutate(gs = paste(Genus,Species)) %>% 
  group_by(gs) %>% 
  summarise(n())

df_ms <- bc %>% 
  filter(Genus == "Mazzaella", Species == "splendens")


leaflet() %>% 
  addProviderTiles("OpenStreetMap.Mapnik") %>% 
  addCircleMarkers(lat = df_ms$Geo_LatDecimal, lng = df_ms$Geo_LongDecimal,
                   popup = paste("Collector:", df_ms$`Primary Collector`,
                                 "Date collected:", df_ms$`Date Collected`,
                                 ", Location:", df_ms$Location, 
                                 ", Collect no.:",df_ms$`Collector Number`),
                   clusterOptions = markerClusterOptions())

```

#### BONUS - WMS tiles
- **W**eb **M**ap tile **S**ervice: *standard protocol for serving pre-rendered or run-time computed georeferenced map tiles over the Internet*
- Get this from [DataBC](https://catalogue.data.gov.bc.ca/)
- 2000 + data sets

##### What you will need to plot WMS tiles:
- baseUrl
- layers


- We will use [Parks and Protected Areas](https://catalogue.data.gov.bc.ca/dataset/parks-and-protected-areas-section-boundaries)
- [click](https://catalogue.data.gov.bc.ca/dataset/parks-and-protected-areas-section-boundaries/resource/bb167646-e401-4a2a-bb3d-b7a44c11c097) WMS getCapabilities request option
- copy the [url](`https://openmaps.gov.bc.ca/geo/pub/WHSE_ENVIRONMENTAL_MONITORING.SHZN_HAB_OBS_POLYS_SV/ows?service=WMS`), removing `&request=GetCapabilities`

###### Layers
[Link to file](https://openmaps.gov.bc.ca/geo/pub/WHSE_ADMIN_BOUNDARIES.ADM_BC_PARKS_SECTIONS_SP/ows?service=WMS&request=GetCapabilities)
- search for `pub`
- should find something like this `pub:WHSE_ADMIN_BOUNDARIES.ADM_BC_PARKS_SECTIONS_SP`

###### Options
- add `WMSTileOptions(transparent = TRUE, zoomToBoundsOnCLick = TRUE)`

*Note * You may need to zoom in to see the layer appear
```{r}
leaflet() %>% 
  addProviderTiles("OpenStreetMap.Mapnik") %>% 
  addWMSTiles("https://openmaps.gov.bc.ca/geo/pub/WHSE_ADMIN_BOUNDARIES.ADM_BC_PARKS_SECTIONS_SP/ows?service=WMS",
              layers = "pub:WHSE_ADMIN_BOUNDARIES.ADM_BC_PARKS_SECTIONS_SP",
              options = WMSTileOptions(transparent = TRUE,
                                       zoomToBoundsOnCLick = TRUE)) %>% 
    addCircleMarkers(lat = df_ms$Geo_LatDecimal, lng = df_ms$Geo_LongDecimal)
```

### Try finding your own wms tiles and plotting
```{r}

```

