---
title: "High quality figures with ggplot2 and beyond"
author: "Travis Blimkie"
date: "January 27th, 2021"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  comment = ""
)
```

***

## Lesson overview
Today we're going to look at a few ways of improving on default ggplot2
graphics, as well as explore some other packages for plotting data.
Specifically, we're going to look at the `cowplot` package for tweaking our
ggplot2 figures, and the `ggtext` package which allows for more advanced text
formatting on our plots. Then we'll produce interactive graphics via the
`plotly` package.
For this lesson we'll need the built-in `mtcars` data set in addition to the
`gapminder` data to test out these different plotting methods.

## Load packages
Let's start by loading the libraries we'll need.  
```{r libraries}
# install.packages(c("cowplot", "plotly", "ggtext", "tidyverse", "gapminder"))
library(cowplot)
library(plotly)
library(ggtext)
library(gapminder)
library(tidyverse)

data("mtcars")
mtcars <- mtcars %>% rownames_to_column("car")
```

<br>

## Our first plot & themes
Let's start by making a standard plot using `ggplot2` to get our baseline. 
```{r basic_ggplot}
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point()
```

<br>

Let's see how we can tweak this plot using the available methods. We'll be
making use of two functions here: `labs()` to create labels, and `theme()` to
customize the appearance of our plot. Let's change the axis labels, add a title,
and change the overall theme with `theme_bw()`.
```{r tweak_ggplot_1}
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point() +
  labs(
    x = "Number of Cylinders", 
    y = "Miles per Gallon", 
    title = "Comparing Engine Size and Fuel Efficiency"
  ) +
  theme_bw() +
  theme(axis.ticks = element_blank())
```

<br>

The `theme_bw()` function is acting as a "preset", changing multiple arguments
of the regular `theme()` function at once. There are some additional versions of
`theme_bw()`, such as `theme_classic()`, and `theme_grey()` (the default of
ggplot2). For `theme()` itself, there are a whole range of arguments that can be
used to alter the appearance of the plot. Any argument you want to specify must
be in the format of `element_x()` (like the `element_blank()` we used earlier).
Let's build on this example by increasing the font size, as the default is often
too small.
```{r tweak_ggplot_2}
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point(size = 3, pch = 21, fill = "dodgerblue4", colour = "black") +
  labs(
    x = "Number of Cylinders", 
    y = "Miles per Gallon", 
    title = "Comparing Engine Size and Fuel Efficiency"
  ) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    axis.text = element_text(size = 12, colour = "black"),
    title = element_text(size = 14, colour = "dodgerblue4"),
    panel.grid.minor = element_blank()
  )
```

<br>

Now our plot looks a lot nicer!  

<br>

## Making use of cowplot
Another alternative to making a bunch of little changes ourselves is to use
another package, `cowplot`. This package comes with suite of functions, but
let's start with `theme_cowplot()`, which makes a variety of changes to the
default look of ggplot2 graphics.
```{r cowplot_1}
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point(size = 3, pch = 21, fill = "dodgerblue4", colour = "black") +
  labs(
    x = "Number of Cylinders", 
    y = "Miles per Gallon", 
    title = "Comparing Engine Size and Fuel Efficiency"
  ) +
  theme_cowplot()
```

<br>

Another really useful feature of the `cowplot` package is the ability to combine
multiple plots into a multi-panel style figure using the `plot_grid()` function.
Let's test it out with an example.
```{r plot_grid, fig.height = 8}
mpg_cyl_plot <- ggplot(mtcars, aes(cyl, mpg)) +
  geom_point(size = 3, pch = 21, fill = "dodgerblue4", colour = "black") +
  labs(
    x = "Number of Cylinders", 
    y = "Miles per Gallon", 
    title = "Comparing Engine Size and Fuel Efficiency"
  ) +
  theme_cowplot()

disp_hp_plot <- ggplot(mtcars, aes(disp, hp)) +
  geom_smooth(method = "lm", colour = "grey") +
  geom_point(size = 3, pch = 21, fill = "tomato", colour = "black") +
  labs(
    x = "Displacement", 
    y = "Horse Power", 
    title = "Comparing Engine Size and Displacement"
  ) +
  theme_cowplot()

plot_grid(
  mpg_cyl_plot, disp_hp_plot, 
  nrow = 2, ncol = 1, align = "h", labels = "AUTO")
```

<br>

## Customizing text with ggtext
The next package we'll use, `ggtext`, allows us to format any text on our plots,
using a bit of HTML. Let's start with an example.
```{r ggtext_1}
gapminder_filtered <- gapminder %>% 
  filter(country %in% c("Canada", "Germany", "France", "Australia")) %>% 
  mutate(country_continent = paste0(country, "<br><i>(", continent, ")</i>"))

ggplot(gapminder_filtered, aes(country_continent, pop, fill = country)) +
  geom_boxplot() +
  theme_cowplot() +
  theme(
    axis.text.x = element_markdown(),
    legend.position = "none"
  ) +
  labs(x = "Country", y = "Population")

```

<br>

## Interactive plots with plotly
Now let's move on to another plotting library, `plotly`. Unlike ggplot2, this
package allows us to make interactive HTML plots. Let's see what this looks like
with an example. Note the syntax is a bit different then ggplot2, so pay close
attention to how we set up our plot.
```{r plotly_1}
plot_ly(mtcars, x = ~cyl, y = ~mpg, type = "scatter", text = ~car) %>% 
  plotly::layout(
    title = "Comparing Engine Size and Fuel Efficiency",
    xaxis = list(title = "Number of Cylinders"),
    yaxis = list(title = "Miles per Gallons")
  )
```

<br>

As you can see, hovering our cursor over the points displays the coordinates as well as the name of the vehicle.  
Let's see how we can tweak this some more. 
```{r plotly_2}
fit <- lm(mpg ~ cyl, data = mtcars)

plot_ly(mtcars, x = ~cyl, y = ~mpg, type = "scatter", text = ~car) %>% 
  add_lines(x = ~cyl, y = fitted(fit), line = list(color = "black")) %>% 
  plotly::layout(
    title = "Comparing Engine Size and Fuel Efficiency",
    xaxis = list(title = "Number of Cylinders"),
    yaxis = list(title = "Miles per Gallons")
  )
```

<br>

We can also extensively format the hover text displayed on the plot, using a bit
of HTML language.
```{r plotly_3}
fit <- lm(mpg ~ cyl, data = mtcars)

plot_ly(mtcars, x = ~cyl) %>%
  add_trace(
    y = ~mpg, 
    mode = "markers", 
    marker = list(size = 10),
    hoverinfo = "text",
    text = ~paste0("<b>Car: </b>", car)
  ) %>% 
  add_trace(y = fitted(fit), mode = "lines") %>% 
  plotly::layout(
    title = "Comparing Engine Size and Fuel Efficiency",
    xaxis = list(title = "Number of Cylinders"),
    yaxis = list(title = "Miles per Gallons")
  )
```

<br>

Making a boxplot using plotly, specifying custom colours, and adding lines along
the x and y axis.
```{r plotly_4}
plot_ly(mtcars, x = ~cyl) %>%
  add_boxplot(
    y = ~mpg, 
    marker = list(color = "black"),
    line = list(color = "black"), 
    fillcolor = "white"
  ) %>% 
  plotly::layout(
    title = "Comparing Engine Size and Fuel Efficiency",
    xaxis = list(showline = TRUE, 
                 title = "Number of Cylinders", 
                 titlefont = list(size = 14), 
                 tickfont = list(size = 14)),
    yaxis = list(showline = TRUE, 
                 title = "Miles per Gallons",
                 titlefont = list(size = 14), 
                 tickfont = list(size = 14))
  )
```
